---
title: "HMDA RD Analysis - Bhutta (2011) Specification"
author: "Bhutta Replication Project"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bhutta (2011) Replication - Modified Specification

This script implements the Bhutta (2011) specification for CRA analysis.

**KEY DIFFERENCES FROM ORIGINAL R CODE:**
1. NO central_city_flag filter (Bhutta does NOT restrict to central city)
2. Uses total housing units >= 100 (not owner-occupied)
3. Includes group quarters filter (< 30%)

## 1. Setup

```{r packages, warning=FALSE, message=FALSE}
library(tidyverse)
library(fixest)
library(modelsummary)
```

```{r filepaths}
# Define paths - adjust as needed
base_path <- file.path(Sys.getenv("DATA_ROOT", "data"), "Inputs/Old/CRA_code/_files2/")
output_path <- file.path(Sys.getenv("OUTPUT_ROOT", "outputs"), "bhutta_replication/output/")

# Create output directory if it doesn't exist
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
```

## 2. Data Loading

This assumes you have the combined_data5 file from the original processing pipeline.
If not, you'll need to run the data preparation steps first.

```{r load_data, eval=FALSE}
# Load the pre-central-city-filter data (step 15)
# This is the data BEFORE the central city filter is applied
combined_data5 <- read.csv(paste0(base_path, "1990to2006/out/combined_lar_step15.csv"))
```

## 3. Apply Bhutta Specification Filters

**CRITICAL: This is where we differ from the original R code**

```{r bhutta_filters, eval=FALSE}
# ============================================================
# BHUTTA SPECIFICATION - KEY DIFFERENCES
# ============================================================

# STEP 1: DO NOT filter to central city
# ORIGINAL CODE (line 657): combined_data6 <- combined_data5 %>% filter(central_city_flag == 1)
# BHUTTA SPEC: Skip this filter entirely
combined_data_bhutta <- combined_data5

# STEP 2: Geographic exclusions (same as original)
combined_data_bhutta <- combined_data_bhutta %>%
    mutate(state_code = state_code.x) %>%
    mutate(county_code = county_code.x) %>%
    filter(state_code != 2) %>% # Exclude Alaska
    filter(state_code != 15) # Exclude Hawaii

# STEP 3: Housing unit filter - USE TOTAL, NOT OWNER-OCCUPIED
# ORIGINAL CODE: filter(ownOcc_housingUnits >= 100)
# BHUTTA SPEC: filter(totalHousingUnits >= 100)
combined_data_bhutta <- combined_data_bhutta %>%
    filter(totalHousingUnits >= 100)

# STEP 4: Group quarters filter (Bhutta uses < 30%)
# This may not be in your data - skip if not available
# combined_data_bhutta <- combined_data_bhutta %>%
#   filter(group_quarters_pct < 0.30)

# Clean up columns
combined_data_bhutta <- combined_data_bhutta %>%
    select(
        -msamd.x,
        -state_code.x, -state_code.y,
        -county_code.x, -county_code.y
    )

cat("Bhutta specification sample size:", nrow(combined_data_bhutta), "\n")
```

## 4. Create Analysis Variables

```{r analysis_vars, eval=FALSE}
# Calculate housing age proportions
combined_data_bhutta$totalHousingUnits <- rowSums(combined_data_bhutta[, c(
    "housingUnits_19891990", "housingUnits_19851988",
    "housingUnits_19801984", "housingUnits_19701979",
    "housingUnits_19601969", "housingUnits_19501959",
    "housingUnits_19401949", "housingUnits_pre1939"
)], na.rm = TRUE)

combined_data_bhutta$prop_pre1940 <- combined_data_bhutta$housingUnits_pre1939 / combined_data_bhutta$totalHousingUnits

combined_data_bhutta$housingUnits_1940_1969 <- combined_data_bhutta$housingUnits_19601969 +
    combined_data_bhutta$housingUnits_19501959 +
    combined_data_bhutta$housingUnits_19401949

combined_data_bhutta$prop_1940_1969 <- combined_data_bhutta$housingUnits_1940_1969 / combined_data_bhutta$totalHousingUnits

# Log transformations
combined_data_bhutta$ln_totalHousingUnits <- log(combined_data_bhutta$totalHousingUnits)
combined_data_bhutta$ln_ownOcc_housingUnits <- log(pmax(combined_data_bhutta$ownOcc_housingUnits, 1))
combined_data_bhutta$ln_ownOcc_housingUnits_medianValue <- log(pmax(combined_data_bhutta$ownOcc_housingUnits_medianValue, 1))

# Create treatment indicator
combined_data_bhutta$tract_income_low <- ifelse(combined_data_bhutta$decennial_msa_median_family_income_percentage < 0.80, 1, 0)

# Create GSE goal indicator (TM <= 0.90)
combined_data_bhutta$d_mfi_percentage_ninety <- ifelse(combined_data_bhutta$decennial_msa_median_family_income_percentage <= 0.90, 1, 0)

# Grouping variable
combined_data_bhutta <- combined_data_bhutta %>%
    mutate(group_variable = case_when(
        decennial_msa_median_family_income_percentage >= 0.50 & decennial_msa_median_family_income_percentage < 0.75 ~ "g5075",
        decennial_msa_median_family_income_percentage >= 0.75 & decennial_msa_median_family_income_percentage < 0.85 ~ "g7585",
        decennial_msa_median_family_income_percentage > 0.85 & decennial_msa_median_family_income_percentage <= 1.10 ~ "g85110",
        TRUE ~ NA_character_
    ))
```

## 5. Define Control Variables

```{r controls, eval=FALSE}
# Control columns for narrow bandwidth (h=0.05)
control_columns <- c(
    "minority_percentage",
    "ln_ownOcc_housingUnits", # Bhutta uses log owner-occupied
    "ln_ownOcc_housingUnits_medianValue",
    "poverty_level_percentage",
    "prop_pre1940",
    "prop_1940_1969"
)

# Control columns for wide bandwidth (h=0.30) - includes GSE indicator
control_columns30 <- c(
    "minority_percentage",
    "ln_ownOcc_housingUnits",
    "ln_ownOcc_housingUnits_medianValue",
    "poverty_level_percentage",
    "prop_pre1940",
    "prop_1940_1969",
    "d_mfi_percentage_ninety"
)
```

## 6. RD Regression Function

```{r regression_function, eval=FALSE}
run_regression <- function(data, outcome_var, h, control_function) {
    # Select the appropriate control columns based on bandwidth
    if (h == 0.30) {
        control_vars <- control_columns30
    } else {
        control_vars <- control_columns
    }

    # Create the formula based on the control function
    if (control_function == "linear") {
        formula <- as.formula(paste0(
            outcome_var, " ~ tract_income_low + decennial_msa_median_family_income_percentage + ",
            paste(control_vars, collapse = " + "), " | metropolitan_area_code"
        ))
    } else if (control_function == "cubic") {
        formula <- as.formula(paste0(
            outcome_var, " ~ tract_income_low + poly(decennial_msa_median_family_income_percentage, 3) + ",
            paste(control_vars, collapse = " + "), " | metropolitan_area_code"
        ))
    } else {
        stop("Invalid control function specified.")
    }

    # Filter data based on bandwidth (h)
    data_filtered <- data %>%
        filter(abs(decennial_msa_median_family_income_percentage - 0.80) <= h)

    # Run the regression using fixest package for clustered standard errors
    model <- feols(formula, data = data_filtered, cluster = ~metropolitan_area_code)

    return(model)
}
```

## 7. Run Regressions

```{r run_regressions, eval=FALSE}
# Run regressions for ALL MSAs (Bhutta specification)
cat("\n===== BHUTTA SPECIFICATION RESULTS =====\n")
cat("(NO central city filter, total HU >= 100)\n\n")

model_all_bhutta <- run_regression(combined_data_bhutta, "ln_number_of_loans", h = 0.05, "linear")
model_large_linear_bhutta <- run_regression(combined_data_bhutta_large, "ln_number_of_loans", h = 0.05, "linear")
model_large_cubic_bhutta <- run_regression(combined_data_bhutta_large, "ln_number_of_loans", h = 0.30, "cubic")

# Output results
modelsummary(
    list(
        "All MSAs" = model_all_bhutta,
        "Large MSAs (Linear)" = model_large_linear_bhutta,
        "Large MSAs (Cubic)" = model_large_cubic_bhutta
    ),
    output = paste0(output_path, "bhutta_spec_regression_results.xlsx"),
    title = "Bhutta (2011) Specification - RD Results",
    stars = TRUE,
    gof_omit = "DF|Deviance|Log.Lik.|AIC|BIC"
)
```

## 8. Sample Size Comparison

```{r sample_comparison, eval=FALSE}
# Compare sample sizes between original and Bhutta specification
cat("\n===== SAMPLE SIZE COMPARISON =====\n\n")

# Original specification (with central city filter)
original_sample <- combined_data5 %>%
    filter(central_city_flag == 1) %>%
    filter(state_code.x != 2 & state_code.x != 15) %>%
    filter(ownOcc_housingUnits >= 100)

# Bhutta specification (without central city filter)
bhutta_sample <- combined_data5 %>%
    filter(state_code.x != 2 & state_code.x != 15) %>%
    filter(totalHousingUnits >= 100)

cat("Original specification (central city only):", nrow(original_sample), "\n")
cat("Bhutta specification (all MSA tracts):     ", nrow(bhutta_sample), "\n")
cat("Difference:                                ", nrow(bhutta_sample) - nrow(original_sample), "\n")
cat(
    "Percent removed by central city filter:    ",
    round((nrow(bhutta_sample) - nrow(original_sample)) / nrow(bhutta_sample) * 100, 1), "%\n"
)

# Narrow bandwidth comparison
original_narrow <- original_sample %>%
    filter(abs(decennial_msa_median_family_income_percentage - 0.80) <= 0.05)
bhutta_narrow <- bhutta_sample %>%
    filter(abs(decennial_msa_median_family_income_percentage - 0.80) <= 0.05)

cat("\nNarrow bandwidth (h=0.05):\n")
cat("Original:     ", nrow(original_narrow), "\n")
cat("Bhutta spec:  ", nrow(bhutta_narrow), "\n")
```

## 9. Expected Results

Based on our Python analysis:

| Specification | β Coefficient | SE | Sign |
|--------------|---------------|-----|------|
| Original R (central city) | -0.088 | (0.086) | ❌ Wrong |
| Bhutta spec (all MSA) | +0.007 to +0.08 | (~0.03) | ✅ Correct |
| Bhutta (2011) paper | +0.0764*** | (0.0274) | ✅ Target |

The key insight is that removing the central city filter should:
1. **Reverse the sign** from negative to positive
2. **Reduce standard errors** (larger sample)
3. **Increase statistical significance**

---

*This script implements the Bhutta (2011) specification without modifying the original R code.*
*See SPECIFICATION_COMPARISON_REPORT.md for full analysis.*
